Emulsions Stabilized by Magnetic Ellipsoidal Particles: A Lattice Boltzmann Study

### 2025-04-01 chapters/dissertation-tap.tex
Changes 1. Refactoring LBM methods explanation 2. Removing contact angle plots for model validation 3. Changes psi vs t figure in paper 1
Rewrote abstract

% Title and author are also used for PDF file properties
% No special character or commands can be used for the PDF definition; 
% use the [options] paramater to specify a different title or author 
% to remove special characters or commands like \\ for example.
\title[First Line of Title Second Line of Title]{Lattice Boltzmann simulations of magnetically responsive\\
ellipsoidal particle stabilized emulsions}
\author{Nikhil Karthikeyan}
\type{dissertation}
\degree{Doctor of Philosophy}
\majorfieldtrue\majorfield{Materials Science and Engineering}
\degreedate{Spring 2025}
% Optional PDF properties
\keywords{Keyord,Keyword,Keyword}
\subject{Subject}

@@

\prefacesectiontoc{Abstract}
%\input{abstract} % This file (abstract.tex) contains the text
                 % for an abstract or type text here.
    Porous materials are important in a wide range of applications, including water filtration, 
    catalyst supports, battery electrodes, and bioengineered materials. To accommodate a broader range of pore 
    sizes and enable scalable fabrication with minimal waste, bottom-up synthesis techniques have gained 
    increasing attention. Among these, emulsion templating based on bicontinuous interfacially jammed 
    emulsion gels (bijels)—has emerged as a promising approach due to its ability to produce complex, 
    interconnected porous structures.

    Despite these advantages, achieving microstructural control independent of the casting mixture remains a 
    key challenge, particularly in continuous bijel fabrication methods such as Solvent Transfer Induced Phase 
    Separation (STrIPS). Additionally, in situ control strategies for tailoring properties such 
    as separation efficiency can be useful additions to control material performance. 
    Stimuli-responsive approaches have been identified as a potential route to 
    overcome these constraints. This study focuses on the magnetic stimuli-response of bijels stabilized by 
    magnetically responsive ellipsoidal microparticles, examining both the resulting microstructural 
    modifications and their rheological implications.

    It is shown that the application of a magnetic field during bijel formation induces microstructural 
    changes in systems stabilized by ellipsoidal particles—effects not observed in bijels stabilized by 
    spherical particles. The applied field causes particle reorientation along its direction, leading to 
    directional jamming and the development of anisotropic microstructures.

    Post-synthesis, bijels stabilized by ellipsoidal particles exhibit tunable microstructure. This responsiveness 
    is governed by the degree of particle order present prior to field application, as well as the extent of 
    reordering induced by the field. The system exhibits state-dependent behavior, indicating that the timing of 
    magnetic field application significantly influences the magnitude of the microstructural response characterized.

    Rheological characterization reveals that bijels retain their shear-thinning behavior under applied magnetic 
    fields. However, the degree of particle ordering is found to influence both the evolution of the microstructure 
    under shear and the corresponding rheological response, highlighting the interplay between field-induced 
    structuring and mechanical performance.
\end{front}

>>>>

% Title and author are also used for PDF file properties
% No special character or commands can be used for the PDF definition; 
% use the [options] paramater to specify a different title or author 
% to remove special characters or commands like \\ for example.
\title[First Line of Title Second Line of Title]{Emulsions Stabilized by Magnetic Ellipsoidal Particles: A Lattice Boltzmann Study}
\author{Nikhil Karthikeyan}
\type{dissertation}
\degree{Doctor of Philosophy}
\majorfieldtrue\majorfield{Materials Science and Engineering}
\degreedate{Spring 2025}
% Optional PDF properties
\keywords{Keyord,Keyword,Keyword}
\subject{Subject}

@@

\prefacesectiontoc{Abstract}
%\input{abstract} % This file (abstract.tex) contains the text
                 % for an abstract or type text here.
    Porous materials are important across a wide range of applications, including water filtration, catalyst supports, 
    battery electrodes, and bioengineered materials. To accommodate a broader range of pore sizes and enable scalable 
    fabrication with minimal waste, bottom-up synthesis techniques have gained increasing attention. Emulsion templating 
    leverages thermodynamically or kinetically arrested structures formed by phase-separating fluids. Among these, bicontinuous 
    interfacially jammed emulsion gels (bijels) are particularly promising due to their tortuous and co-continuous microstructures.

    Bijels form via thermally induced spinodal decomposition of partially miscible fluid mixtures in the presence of neutrally 
    wetting particles. As the fluid domains coarsen, particles adsorb onto the interface until the interfacial area matches the 
    total cross-sectional area of the particles, resulting in a jammed monolayer that locks the microstructure in place. Traditional 
    bijel synthesis via thermal phase separation is not readily scalable. However, newer techniques such as Solvent Transfer Induced 
    Phase Separation (STrIPS) have been developed to address this limitation, utilizing the removal of solvent from a bijel casting
    mixture to induce phase separation. However, the microstructure obtained is coupled to the casting mixture composition, 
    and the flow rate during STrIPS.

    One promising approach to modulate bijel microstructure is through stimulus-responsive systems. Magnetic stimuli, in particular, 
    offer a non-invasive, controllable, and targeted mechanism. Earlier studies using spherical particles under magnetic fields showed 
    limited microstructural changes. More recent work, however, has revealed that ellipsoidal particles can respond to magnetic actuation 
    by tilting, leading to the formation of particle chains or rings at interfaces. This phenomenon opens the door to controllable 
    modifications in bijel microstructures using magnetic fields.

    This dissertation investigates whether constant magnetic fields can be used to control the microstructure of bijels stabilized by 
    magnetically responsive ellipsoidal particles. To explore this, we employ a multicomponent Lattice Boltzmann Method coupled with a 
    molecular dynamics representation of rigid particles and a dipole-based magnetic field model. The system simulates a water-2,6-lutidine 
    mixture stabilized by micron-sized nickel-coated polystyrene particles.

    We first examine the ability of magnetic fields to influence bijel formation by applying a constant field during spinodal decomposition 
    of fluid mixtures containing disc-like, spherical, and rod-like particles. While no significant change in domain size is observed for 
    spherical particles, bijels stabilized by discs and rods show slight increases in domain size and become anisotropic, as measured by 
    tortuosity and directional domain size.Further analysis reveals that the coarsening rates become direction-specific for ellipsoidal particles, 
    suggesting that jamming occurs anisotropically due to particle alignment with the magnetic field. This alignment, governed by the 
    orientation of the magnetic moment relative to the particle's long axis, also affects how particles arrange themselves at the fluid 
    interface.

    Having established that magnetic fields can influence microstructure during synthesis, we then assess whether similar control can be exerted 
    after bijel formation. This has implications for applications like crossflow reactors and filtration systems, where reversible control over 
    permeability and flow resistance is desirable. By incrementally increasing and decreasing the magnetic field on a bijel stabilized with 
    rod-like particles, we observe that the microstructure can indeed be modified post-synthesis. Domain size increases nonlinearly with 
    applied field strength and remains altered even after the field is reduced, indicating a saturation threshold and a history-dependent 
    response.

    To probe this further, we evaluate field-driven coarsening in bijels stabilized with rod and disc-like particles. Upon field application, 
    particles reorient, leading to anisotropic domains similar to those formed during synthesis under a field. The temporal evolution of 
    microstructure involves increased particle ordering, interface alignment, and rearrangement, demonstrating a complex interplay of factors. 
    Notably, the extent of domain size change is negatively correlated with the initial ordering of particles, suggesting a memory effect 
    rooted in the starting microstructure.

    Given that many magnetically responsive materials exhibit field-dependent rheological behavior (e.g., shear thickening in ferrofluids), 
    we investigate whether bijels exhibit similar effects. Prior studies have shown that shear can induce domain coarsening and particle 
    ejection in bijels. Additionally, emulsions stabilized by ellipsoidal particles show viscosity decreases with increased ordering due 
    to weakened particle-fluid interactions. In this work, we explore how magnetic fields and initial microstructure affect the shear 
    response of bijels stabilized by ellipsoidal particles.As particle ordering increases, the viscosity and shear-thinning behavior decrease for 
    bijels stabilized by disc- and rod-like particles. Interestingly, the yield stress decreases with increased order for discs, but increases for 
    rods due to differences in local particle arrangement and frictional contacts. Under shear, domain coarsening occurs due to shear induced particle 
    tilting out of the interface.

    In conclusion, this dissertation demonstrates that magnetic fields offer a viable and tunable method for controlling bijel microstructures both 
    during and after synthesis, and it provides a detailed characterization of the constant shear response of magnetically responsive bijels. These findings 
    lay the groundwork for developing adaptive, magnetically responsive porous materials through both conventional synthesis techniques and Solvent 
    Transfer Induced Phase Separation (STrIPS), enabling scalable and application-specific design.

\end{front}
